To be FAIR: Theory Specification Needs an Update

### 2025-06-20 manuscript.Rmd
Update manuscript.Rmd
R-package -> R package

## The Need for FAIR theory

@@

The replication crisis has brought the inadequacies of contemporary theoretical practices in psychology and other fields into focus.
Psychological theories often fall short of the FAIR principles:
they are hard to find and access, have no practical uses in scholars' daily workflows beyond providing context for a literature review,
and are more likely to be forgotten or replaced than reused.
These limitations impede cumulative knowledge production in our field,
leading to an accumulation of "one-shot" empirical findings, without commensurate advancement in our principled understanding of psychological phenomena.
We argued that applying the FAIR principles to theory offers a structured solution to these shortcomings.
We demonstrated how to create, version-control, and archive theories as digital information artifacts.
We introduced the `theorytools` R-package to partly automate these processes, reducing the barrier of entry for researchers,
and creating a FAIR resource for theory construction tools and documentation that can be updated as best practices continue to develop.

@@

FAIR theory requires a departure from contemporary practice.
Several factors can expedite such a culture change.
One key factor is the *recognition and rewards* movement:
practices for evaluating scientific output are evolving, with initiatives like the [*Declaration on Research Assessment* (DORA)](https://sfdora.org/read/) and [Coalition for Advancing Research Assessment](https://coara.eu/)
promoting the use of more diverse and meaningful metrics beyond journal impact factors.
Modular publishing capitalizes on these changing metrics,
and FAIR theory allows scholars to be credited for theoretical contributions [@kirczModularityNextForm1998].
Journals that publish theoretical papers could require authors to additionally publish their theories in a FAIR format, cross-referencing the paper,
to expedite its effective reuse and iterative improvement.
A second factor is to lower barriers to adopting FAIR theory by building upon existing widely adopted open science infrastructures.
For this reason, we advocate the use of Git for version control, Zenodo for archiving, and DataCite for standardized metadata.
Barriers to entry can also be lowered by simplifying workflows, which is the goal of the `theorytools` R-package.
Fourth, the availability of Open Educational Materials (OEM) about theory development contributes to doctoral socialization.
These materials allow teachers to incorporate theory development into their curriculum without investing substantial time on course development,
thus educating the next generation to make use of and contribute to FAIR theory.
Finally, community building plays an important role;
the international network of open science communities, reproducibility networks, and other similar initiatives provide platforms for disseminating FAIR theories and related methodological innovations.
Authors can also share their FAIR theories with other early adopters by submitting them to the "FAIR Theory Community" on Zenodo.

@@

Another limitation is the learning curve associated with tools like Git and Zenodo.
The `theorytools` R-package mitigates this limitation for R-users by automating key steps in the process.
Moreover, the initial investment in time can be offset by long-term productivity gains and increased impact of FAIR theory.
One barrier to adopting FAIR theory is cultural resistance to sharing and modifying theories, also known as the "toothbrush problem".
Education might help address this limitation; with this in mind,
we have shared several open educational materials on theory development in the "FAIR Theory Community" on Zenodo, and we encourage others to reuse these and share their materials.

>>>>

## The Need for FAIR Theory

@@

The replication crisis has brought the inadequacies of contemporary theoretical practices in psychology and other fields into focus.
Psychological theories often fall short of the FAIR principles:
they are hard to find and access, have no practical uses in scholars' daily workflows beyond providing context for a literature review,
and are more likely to be forgotten or replaced than reused.
These limitations impede cumulative knowledge production in our field,
leading to an accumulation of "one-shot" empirical findings, without commensurate advancement in our principled understanding of psychological phenomena.
We argued that applying the FAIR principles to theory offers a structured solution to these shortcomings.
We demonstrated how to create, version-control, and archive theories as digital information artifacts.
We introduced the `theorytools` R package to partly automate these processes, reducing the barrier of entry for researchers,
and creating a FAIR resource for theory construction tools and documentation that can be updated as best practices continue to develop.

@@

FAIR theory requires a departure from contemporary practice.
Several factors can expedite such a culture change.
One key factor is the *recognition and rewards* movement:
practices for evaluating scientific output are evolving, with initiatives like the [*Declaration on Research Assessment* (DORA)](https://sfdora.org/read/) and [Coalition for Advancing Research Assessment](https://coara.eu/)
promoting the use of more diverse and meaningful metrics beyond journal impact factors.
Modular publishing capitalizes on these changing metrics,
and FAIR theory allows scholars to be credited for theoretical contributions [@kirczModularityNextForm1998].
Journals that publish theoretical papers could require authors to additionally publish their theories in a FAIR format, cross-referencing the paper,
to expedite its effective reuse and iterative improvement.
A second factor is to lower barriers to adopting FAIR theory by building upon existing widely adopted open science infrastructures.
For this reason, we advocate the use of Git for version control, Zenodo for archiving, and DataCite for standardized metadata.
Barriers to entry can also be lowered by simplifying workflows, which is the goal of the `theorytools` R package.
Fourth, the availability of Open Educational Materials (OEM) about theory development contributes to doctoral socialization.
These materials allow teachers to incorporate theory development into their curriculum without investing substantial time on course development,
thus educating the next generation to make use of and contribute to FAIR theory.
Finally, community building plays an important role;
the international network of open science communities, reproducibility networks, and other similar initiatives provide platforms for disseminating FAIR theories and related methodological innovations.
Authors can also share their FAIR theories with other early adopters by submitting them to the "FAIR Theory Community" on Zenodo.

@@

Another limitation is the learning curve associated with tools like Git and Zenodo.
The `theorytools` R package mitigates this limitation for R-users by automating key steps in the process.
Moreover, the initial investment in time can be offset by long-term productivity gains and increased impact of FAIR theory.
One barrier to adopting FAIR theory is cultural resistance to sharing and modifying theories, also known as the "toothbrush problem".
Education might help address this limitation; with this in mind,
we have shared several open educational materials on theory development in the "FAIR Theory Community" on Zenodo, and we encourage others to reuse these and share their materials.
